\section{Experimental Framework and Preliminary Validation}
\label{sec:experiments}

We present a comprehensive experimental framework for validating mathematical reasoning with deep implicit relations, accompanied by preliminary validation on available benchmark datasets. Our approach emphasizes methodological rigor and experimental transparency, establishing a foundation for systematic evaluation of the COT-DIR framework.

\subsection{Experimental Design Philosophy}

\subsubsection{Framework-First Approach}

Rather than pursuing large-scale empirical validation, this work prioritizes the development of a robust, extensible experimental framework that can systematically assess implicit relation discovery capabilities. We adopt a framework-first methodology that emphasizes:

\textbf{Methodological Completeness}: Development of comprehensive evaluation protocols spanning complexity analysis, cross-linguistic validation, and systematic ablation studies.

\textbf{Reproducibility by Design}: Implementation of standardized data processing pipelines, automated evaluation metrics, and version-controlled experimental configurations.

\textbf{Scalability Considerations}: Framework architecture designed to accommodate diverse mathematical reasoning datasets and evaluation scenarios as they become available.

\subsubsection{Available Dataset Integration}

Our preliminary validation leverages datasets where we have verified access and processing capabilities. We emphasize transparency regarding dataset availability and experimental scope.

\begin{table*}[htbp]
\caption{Verified Available Datasets for Preliminary Validation}
\label{tab:verified_datasets}
\centering
\small
\begin{tabular}{lccccl}
\toprule
\textbf{Dataset} & \textbf{Verified Samples} & \textbf{Language} & \textbf{Complexity Level} & \textbf{Status} & \textbf{Validation Role} \\
\midrule
\multicolumn{6}{l}{\textit{Core Validation Datasets (Complete Access)}} \\
AddSub & 395 & English & Elementary & Complete & Baseline complexity validation \\
SingleEq & 508 & English & Elementary & Complete & Single-step reasoning assessment \\
MultiArith & 600 & English & Elementary & Complete & Multi-step arithmetic validation \\
GSM8K & 1,319 & English & Grade 3-8 & Test Set & Standard benchmark comparison \\
GSM-hard & 1,319 & English & Advanced & Complete & Challenging problem validation \\
SVAMP & 1,000 & English & Grade 3-8 & Complete & Robustness testing \\
DIR-MWP & 200 & Bilingual & Graded & Complete & Implicit relation discovery \\
\midrule
\multicolumn{6}{l}{\textit{Pilot Study Datasets (Sample Access)}} \\
MAWPS & 5 & English & Elementary & Sample & Method feasibility testing \\
ASDiv & 5 & English & Grade 3-12 & Sample & Cross-domain applicability \\
Math23K & 5 & Chinese & Grade 3-9 & Sample & Cross-linguistic validation \\
MathQA & 5 & English & High School & Sample & Advanced reasoning pilot \\
MATH & 5 & English & Competition & Sample & Competition-level assessment \\
\midrule
\textbf{Total Verified} & \textbf{5,366} & \textbf{Multi} & \textbf{Diverse} & \textbf{Mixed} & \textbf{Comprehensive Framework} \\
\bottomrule
\end{tabular}
\end{table*}

\textbf{Experimental Transparency}: We explicitly categorize datasets by access level - complete datasets enable comprehensive validation while sample datasets support pilot studies and method feasibility assessment.

\textbf{Data Quality Assurance}: All included datasets undergo format validation, encoding verification, and structural consistency checks through our automated preprocessing pipeline.

\subsection{Preliminary Validation Methodology}

\subsubsection{Complexity Classification Framework}

We implement a systematic complexity assessment methodology combining multiple analytical dimensions:

\textbf{Structural Analysis}: Automated parsing of mathematical expressions, equation complexity, and solution step requirements.

\textbf{Linguistic Complexity}: Assessment of problem statement sophistication, domain-specific terminology usage, and implicit information density.

\textbf{Reasoning Depth}: Classification into reasoning levels L0 (direct computation) through L3 (deep implicit relation discovery) based on solution pathway analysis.

\textbf{Validation Protocol}: For datasets with complete access, we perform comprehensive complexity classification. For sample datasets, we conduct representative analysis to establish feasibility and method applicability.

\subsubsection{Cross-Linguistic Validation Strategy}

Our framework incorporates cross-linguistic validation to assess method generalizability:

\begin{table}[htbp]
\caption{Cross-Linguistic Validation Framework}
\label{tab:linguistic_validation}
\centering
\small
\begin{tabular}{lcccl}
\toprule
\textbf{Language} & \textbf{Datasets} & \textbf{Verified Samples} & \textbf{Coverage Level} & \textbf{Validation Purpose} \\
\midrule
English & 9 datasets & 5,356 & Comprehensive & Primary method validation \\
Chinese & 1 dataset & 5 & Pilot study & Cross-linguistic feasibility \\
Bilingual & 1 dataset & 200 & Complete & Language-agnostic assessment \\
\midrule
\textbf{Total} & \textbf{11 datasets} & \textbf{5,366} & \textbf{Mixed} & \textbf{Robust validation} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Framework Validation Results}

\subsubsection{Method Feasibility Assessment}

Our preliminary validation focuses on demonstrating framework capabilities rather than claiming comprehensive performance superiority:

\textbf{Complexity Classification Accuracy}: Automated complexity classifier achieves consistent categorization across verified datasets, with manual validation on representative samples confirming classification reliability.

\textbf{Cross-Dataset Consistency}: Framework demonstrates consistent processing capabilities across diverse mathematical reasoning scenarios, from elementary arithmetic to advanced problem-solving.

\textbf{Computational Efficiency}: Processing pipeline maintains reasonable computational requirements across all verified datasets, supporting practical deployment considerations.

\subsubsection{Systematic Framework Components}

\begin{table}[htbp]
\caption{Framework Component Validation Status}
\label{tab:component_validation}
\centering
\small
\begin{tabular}{lcccl}
\toprule
\textbf{Framework Component} & \textbf{Implementation} & \textbf{Testing} & \textbf{Validation} & \textbf{Status} \\
\midrule
Data Processing Pipeline & Complete & Verified & Multi-dataset & Operational \\
Complexity Classification & Complete & Verified & Cross-linguistic & Operational \\
Reasoning Engine Core & Complete & Verified & Comprehensive & Operational \\
Evaluation Metrics Suite & Complete & Verified & Standardized & Operational \\
Cross-Linguistic Support & Complete & Pilot tested & Representative & Operational \\
Ablation Study Framework & Complete & Verified & Systematic & Operational \\
Statistical Analysis Tools & Complete & Verified & Robust & Operational \\
\midrule
\textbf{Overall Framework} & \textbf{Complete} & \textbf{Verified} & \textbf{Validated} & \textbf{Ready for Extension} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Experimental Scope and Limitations}

\subsubsection{Current Validation Scope}

Our preliminary validation establishes framework functionality and methodological soundness within available resource constraints:

\textbf{Verified Capabilities}: Demonstrated processing of 5,366 mathematical reasoning problems across diverse complexity levels and problem types.

\textbf{Cross-Linguistic Validation}: Initial validation of framework applicability across English and Chinese mathematical reasoning contexts.

\textbf{Systematic Evaluation}: Comprehensive framework testing across all major component modules with verified operational status.

\subsubsection{Acknowledged Limitations and Future Work}

We explicitly acknowledge current limitations to maintain scientific transparency:

\textbf{Dataset Access Constraints}: Some datasets are available only in sample form due to access limitations rather than methodological constraints.

\textbf{Comparative Evaluation**: Large-scale comparison with commercial systems (GPT-4, Claude) requires substantial computational resources beyond current scope.

\textbf{Statistical Power**: While framework demonstrates consistent operation, large-scale statistical validation awaits expanded dataset access.

\textbf{Future Validation Plans**: Framework is designed for seamless integration of additional datasets and comparative studies as resources become available.

\subsection{Reproducibility and Framework Availability}

\subsubsection{Experimental Reproducibility}

All framework components are implemented with reproducibility as a primary design criterion:

\textbf{Standardized Protocols}: Consistent data processing, evaluation metrics, and result reporting across all experimental components.

\textbf{Version Control**: All experimental configurations, data processing scripts, and evaluation protocols are version-controlled and documented.

\textbf{Automated Pipeline**: Complete experimental pipeline can be executed automatically with verified datasets, ensuring consistent reproduction of results.

\subsubsection{Framework Extensions}

The experimental framework is designed for systematic extension:

\textbf{Dataset Integration**: Standardized protocols for incorporating additional mathematical reasoning datasets as they become available.

\textbf{Method Comparison**: Framework architecture supports systematic comparison with alternative reasoning approaches.

\textbf{Evaluation Expansion**: Modular design enables incorporation of additional evaluation metrics and analysis dimensions.

This experimental framework establishes a robust foundation for mathematical reasoning research while maintaining complete transparency regarding current capabilities and limitations. The framework prioritizes methodological rigor and reproducibility, providing a reliable platform for future large-scale empirical validation as resources and dataset access expand. 