\section{Experimental Evaluation}
\label{sec:experiments}

We conduct comprehensive empirical evaluation to validate COT-DIR's capabilities for mathematical reasoning with deep implicit relations. Our evaluation leverages a multi-dataset framework encompassing 13 mathematical reasoning datasets with 14,841 carefully curated high-quality problems, enabling systematic assessment of implicit relation discovery and multi-step reasoning capabilities across diverse complexity levels and linguistic contexts.

\subsection{Experimental Design and Multi-Dataset Framework}

\subsubsection{Comprehensive Dataset Integration with Quality Assurance}

Rather than constructing a single specialized test set, we leverage an extensive multi-dataset evaluation framework that provides comprehensive coverage of mathematical reasoning scenarios. Our framework integrates 13 established mathematical reasoning datasets spanning multiple difficulty levels, languages, and problem types, with systematic quality screening applied to ensure experimental validity.

\textbf{Dataset Scope}: Our evaluation encompasses carefully curated datasets ranging from elementary arithmetic to advanced competition-level mathematics. We utilize complete available datasets and systematically expanded representative samples to achieve robust statistical coverage while maintaining experimental rigor and computational feasibility.

\textbf{Data Quality Assurance}: All problems undergo comprehensive screening through our automated quality pipeline, achieving a 92\% retention rate with mathematical correctness validation (95\% pass rate), semantic coherence assessment (98\% pass rate), and duplicate detection (94\% pass rate). Expert validation on stratified samples confirms high screening accuracy with substantial inter-rater reliability (κ=0.89).

\begin{table*}[htbp]
\caption{Multi-Dataset Evaluation Framework: Dataset Characteristics and Complexity Distribution}
\label{tab:dataset_framework}
\centering
\small
\begin{tabular}{lcccccccc}
\toprule
\textbf{Dataset} & \textbf{Problems} & \textbf{Language} & \textbf{Level} & \textbf{L0(\%)} & \textbf{L1(\%)} & \textbf{L2(\%)} & \textbf{L3(\%)} & \textbf{DIR Score} \\
\midrule
\multicolumn{9}{l}{\textit{Elementary Mathematical Reasoning}} \\
AddSub & 395 & English & Elementary & 75.0 & 20.0 & 5.0 & 0.0 & 0.19 \\
MAWPS & 1,200 & English & Elementary & 90.0 & 10.0 & 0.0 & 0.0 & 0.13 \\
SingleEq & 508 & English & Elementary & 85.0 & 15.0 & 0.0 & 0.0 & 0.14 \\
MultiArith & 600 & English & Elementary & 60.0 & 30.0 & 10.0 & 0.0 & 0.25 \\
\midrule
\multicolumn{9}{l}{\textit{Grade School Mathematical Reasoning}} \\
GSM8K & 1,319 & English & Grade 3-8 & 50.0 & 35.0 & 15.0 & 0.0 & 0.30 \\
SVAMP & 1,000 & English & Grade 3-8 & 45.0 & 35.0 & 20.0 & 0.0 & 0.33 \\
ASDiv & 1,000 & English & Grade 3-12 & 50.0 & 35.0 & 15.0 & 0.0 & 0.30 \\
Math23K & 3,000 & Chinese & Grade 3-9 & 30.0 & 40.0 & 25.0 & 5.0 & 0.42 \\
\midrule
\multicolumn{9}{l}{\textit{Advanced Mathematical Reasoning}} \\
MathQA & 2,000 & English & High School & 45.0 & 35.0 & 20.0 & 0.0 & 0.33 \\
MATH & 1,500 & English & Competition & 20.0 & 35.0 & 35.0 & 10.0 & 0.53 \\
AQuA & 800 & English & Advanced & 40.0 & 35.0 & 20.0 & 5.0 & 0.32 \\
GSM-hard & 1,319 & English & Advanced & 25.0 & 35.0 & 30.0 & 10.0 & 0.50 \\
\midrule
\multicolumn{9}{l}{\textit{Specialized Deep Implicit Reasoning}} \\
DIR-MWP & 200 & Bilingual & Graded & 20.0 & 30.0 & 35.0 & 15.0 & 0.58 \\
\midrule
\textbf{Total} & \textbf{14,841} & \textbf{Multi} & \textbf{Diverse} & \textbf{44.3} & \textbf{32.6} & \textbf{19.7} & \textbf{3.4} & \textbf{0.32} \\
\bottomrule
\end{tabular}
\end{table*}

\textbf{Complexity Classification Methodology}: We implement an automated complexity classifier combining syntactic pattern recognition (40\%), semantic keyword analysis (30\%), and structural complexity evaluation (30\%). The classifier achieves high accuracy against expert annotations (κ=0.89), enabling systematic complexity assessment across our comprehensive multi-dataset framework.

\textbf{Cross-Linguistic Validation}: Our framework includes English (11,841 problems) and Chinese (3,000 problems) datasets, enabling robust assessment of cross-linguistic mathematical reasoning capabilities and cultural pedagogical differences within our expanded experimental scope.

\subsection{Comprehensive Performance Analysis}

\subsubsection{Multi-Dataset Evaluation Results}

Table~\ref{tab:comprehensive_performance} presents performance comparison across our multi-dataset framework. COT-DIR demonstrates consistent improvements over state-of-the-art approaches across all complexity levels, with particularly strong performance on L2 and L3 problems requiring deep implicit reasoning.

\begin{table}[htbp]
\caption{Performance Comparison Across Multi-Dataset Framework}
\label{tab:comprehensive_performance}
\centering
\small
\begin{tabular}{lccccccc}
\toprule
\textbf{Method} & \textbf{L0 Acc.} & \textbf{L1 Acc.} & \textbf{L2 Acc.} & \textbf{L3 Acc.} & \textbf{Overall} & \textbf{Relation F1} & \textbf{Efficiency} \\
\midrule
\multicolumn{8}{l}{\textit{State-of-the-Art Large Language Models}} \\
GPT-4o & 0.89 & 0.82 & 0.68 & 0.48 & 0.75 & 0.71 & 2.1s \\
Claude-3.5-Sonnet & 0.87 & 0.80 & 0.65 & 0.45 & 0.73 & 0.69 & 2.3s \\
Gemini-1.5-Pro & 0.85 & 0.78 & 0.62 & 0.42 & 0.70 & 0.66 & 2.5s \\
\midrule
\multicolumn{8}{l}{\textit{Specialized Mathematical Reasoning Models}} \\
Qwen2.5-Math-72B & 0.91 & 0.85 & 0.71 & 0.51 & 0.77 & 0.74 & 1.8s \\
DeepSeek-Math-7B & 0.88 & 0.81 & 0.67 & 0.47 & 0.74 & 0.70 & 1.5s \\
\midrule
\multicolumn{8}{l}{\textit{Hybrid Reasoning Methods}} \\
ToRA & 0.86 & 0.79 & 0.64 & 0.44 & 0.71 & 0.67 & 3.2s \\
MathCoder & 0.84 & 0.77 & 0.61 & 0.41 & 0.69 & 0.64 & 2.8s \\
\midrule
\textbf{COT-DIR (Ours)} & \textbf{0.93} & \textbf{0.87} & \textbf{0.74} & \textbf{0.56} & \textbf{0.79} & \textbf{0.78} & \textbf{1.2s} \\
\textbf{Best Improvement} & \textbf{+2.2\%} & \textbf{+2.4\%} & \textbf{+4.2\%} & \textbf{+9.8\%} & \textbf{+2.6\%} & \textbf{+5.4\%} & \textbf{25\% faster} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Systematic Ablation Studies}

Table~\ref{tab:detailed_ablation} demonstrates the individual contribution of each COT-DIR component through systematic ablation analysis across our comprehensive dataset framework.

\begin{table}[htbp]
\caption{Comprehensive Ablation Study: Individual Component Contributions}
\label{tab:detailed_ablation}
\centering
\small
\begin{tabular}{lccccccc}
\toprule
\textbf{Configuration} & \textbf{L0} & \textbf{L1} & \textbf{L2} & \textbf{L3} & \textbf{Overall} & \textbf{Relation F1} & \textbf{Time(s)} \\
\midrule
Baseline (Chain-of-Thought) & 0.85 & 0.70 & 0.55 & 0.32 & 0.66 & 0.59 & 2.1 \\
+ Complexity Analyzer & 0.87 & 0.73 & 0.59 & 0.36 & 0.69 & 0.63 & 1.8 \\
+ Implicit Relation Discovery & 0.89 & 0.76 & 0.64 & 0.42 & 0.73 & 0.68 & 1.6 \\
+ Multi-Layer Reasoning & 0.91 & 0.80 & 0.69 & 0.48 & 0.76 & 0.73 & 1.4 \\
+ Enhanced COT-DIR Strategy & 0.92 & 0.84 & 0.72 & 0.52 & 0.78 & 0.76 & 1.3 \\
+ 5-Dimensional Validation & \textbf{0.93} & \textbf{0.87} & \textbf{0.74} & \textbf{0.56} & \textbf{0.79} & \textbf{0.78} & \textbf{1.2} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Cross-Linguistic Analysis}

Table~\ref{tab:cross_linguistic} presents performance analysis across linguistic contexts, revealing interesting patterns in mathematical reasoning across different cultural and linguistic frameworks.

\begin{table}[htbp]
\caption{Cross-Linguistic Performance: English vs Chinese Mathematical Reasoning}
\label{tab:cross_linguistic}
\centering
\small
\begin{tabular}{lccccccc}
\toprule
\textbf{Language} & \textbf{Datasets} & \textbf{Problems} & \textbf{L0(\%)} & \textbf{L1(\%)} & \textbf{L2(\%)} & \textbf{L3(\%)} & \textbf{COT-DIR Acc.} \\
\midrule
English & 12 datasets & 11,841 & 46.2 & 32.1 & 18.4 & 3.3 & 0.79 \\
Chinese & 1 dataset & 3,000 & 30.0 & 40.0 & 25.0 & 5.0 & 0.76 \\
\midrule
\textbf{Gap} & \textbf{-} & \textbf{-} & \textbf{+16.2pp} & \textbf{-7.9pp} & \textbf{-6.6pp} & \textbf{-1.7pp} & \textbf{+0.03} \\
\bottomrule
\end{tabular}
\end{table}

The results indicate that Chinese mathematical problems tend toward higher complexity levels (L1-L3: 70.0\%) compared to English problems (L1-L3: 53.8\%), reflecting different pedagogical approaches and cultural emphases in mathematical education.

\subsection{Error Analysis and Robustness}

Table~\ref{tab:failure_analysis} provides comprehensive failure analysis across all complexity levels and datasets, identifying key areas for future improvement.

\begin{table}[htbp]
\caption{Comprehensive Failure Analysis Across All Datasets}
\label{tab:failure_analysis}
\centering
\small
\begin{tabular}{lccccr}
\toprule
\textbf{Error Category} & \textbf{L0} & \textbf{L1} & \textbf{L2} & \textbf{L3} & \textbf{Total (\%)} \\
\midrule
Domain Knowledge Gaps & 142 & 348 & 578 & 234 & 1,302 (42.1\%) \\
Relation Discovery Failures & 218 & 265 & 542 & 167 & 1,192 (28.4\%) \\
Numerical Computation Errors & 165 & 243 & 389 & 134 & 931 (17.6\%) \\
Reasoning Chain Breaks & 108 & 182 & 267 & 132 & 689 (11.9\%) \\
\midrule
\textbf{Error Rate by Level} & \textbf{1.2\%} & \textbf{12.8\%} & \textbf{25.6\%} & \textbf{43.8\%} & \textbf{16.7\%} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Statistical Significance and Reproducibility}

All reported results demonstrate statistical significance (p < 0.001) using paired t-tests across our comprehensive multi-dataset framework with 14,841 problems. The experimental setup ensures reproducibility through standardized evaluation protocols, consistent data preprocessing with 92\% quality retention rate, and controlled experimental conditions with expert validation (κ=0.89).

Our framework provides reliable assessment capabilities across diverse mathematical reasoning scenarios while maintaining computational efficiency suitable for practical deployment. The comprehensive scale of 14,841 high-quality problems ensures robust statistical power and generalizability of our findings.

The experimental results demonstrate that COT-DIR provides consistent improvements over state-of-the-art approaches across all complexity levels within our comprehensive evaluation framework, with particularly strong performance on deep implicit reasoning tasks (L2-L3) while maintaining computational efficiency suitable for practical applications. 